---
title: "Random Forests"
author: "Joe Brown"
date: "2023-03-24"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Random Forests

Attempting to use random forest model to weight models with predicted data against observed data.

Below I build a data frame (`data_t`), which has observed data and predicted data the represent model iterations. The predictaed data strays increasingly further away from the observed data to try and test whether predictions less similar to observed data produce lower weights.

```{r, message=FALSE}
library(tidyverse)
library(randomForest)

# Building df with 4 models that incrementally drift from the observed data
data_t <- tibble(
  observed = c(333, 350, 360, 400, 402, 418, 499, 508, 510, 520),
  run1 = observed + 2,
  run2 = observed + 7,
  run3 = observed + 13,
  run4 = observed + 20)

```

Below I run the forest model with `ntree = 1000` and then use the `IncMSE` values to caluclate a normalized weight for each model. All weight sum to 1.

I then take a few steps to construct a dataframe that is similar to a final hector output - `data_result`.

```{r}
set.seed(seed = 1000)

# Run the random forest model
rf <- randomForest(observed ~ ., data = data_t, ntree = 1000, importance = T)

# Extract the importance values
importance_val <- data.frame(importance(rf))

# Computing weights from the IncMSE values
importance_val$wt <- importance_val$X.IncMSE / sum(importance_val$X.IncMSE)

# gather data
data_long <- gather(data_t[-1])

# Add year values for plotting like Hector data
data_long$year <- rep(1:10, times = 4)

# add run numbers to importance df
importance_val$key <- paste0("run", 1:4) 

# merge the data to the importance_vl df which contains wt
data_result <- merge(data_long, importance_val, by = "key")

```

Finally I plot the results and color code each by their weight produced by the random forest model:

```{r}
plot_test <- ggplot(data = data_result) +
  geom_line(aes(x = year, y = value, 
                group = key,
                color = wt),
            linewidth = 1)+
  scale_color_gradient(high = "dodgerblue4", low = 'lightblue1')+
  theme_light()

# add observed data
data_t$year <- 1:10

plot_test + 
    geom_line(data = data_t, aes(x = year, y = observed),
                color =  'red',
                linewidth = 1)
  
```
____

Same analysis as above, but using hector simulated data:

```{r}
# read in sample hector data in a format that can be used in the rf model 
data <- read.csv("data/new_hector_data.csv")[-1]

# Extract the observed data column and model run columns
set.seed(2)
rf <- randomForest(value_obs ~ ., data = data, ntree = 100000, importance = T)

# importance value extraction from the random forest model
importance_val <- data.frame(importance(rf, type = 1))
# normalizing the importance value so that it can serve as a weight
importance_val$norm <- importance_val$X.IncMSE / sum(importance_val$X.IncMSE)
importance_val$run_number <- 1:nrow(importance_val)

# read in normal hector output and observed data for plotting
hector_data <- read.csv("data/hector_result.csv")[-1]
observed <- read.csv("data/co2_obs.csv")[-1]

# merge hector outpuit with importance values to use as weights
hector_data.res <- merge(hector_data, importance_val, by = "run_number")

# plot the results
plot_test <- ggplot(data = hector_data.res) +
  geom_line(aes(x = year, y = value, 
                group = run_number,
                color = norm),
            linewidth = 1)+
  scale_color_gradient(high = "dodgerblue4", low = 'lightblue1') +
  geom_line(data = observed, aes(x = year, y = co2_ppm),
            color =  'red',
            linewidth = 1) +
  theme_light()
plot_test
```
